RProtoBuf: Efficient Cross-Language Data Serialization in R

### 2013-12-18 papers/rjournal/eddelbuettel-francois-stokely.Rnw
Add basic tables of methods for Messages and Descriptors and improve the basic usage section.
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/rprotobuf@567 edb9625f-4e0d-4859-8d74-9fd3b1da38cb

Modern data collection and analysis pipelines are increasingly being
built using collections of components to better manage software
complexity through reusability, modularity, and fault
isolation \citep{Wegiel:2010:CTT:1932682.1869479}.  Different
programming languages are often used for the different phases of data
analysis -- collection, cleaning, analysis, post-processing, and
presentation in order to take advantage of the unique combination of 
performance, speed of development, and library support offered by
different environments.  Each stage of the data
analysis pipeline may involve storing intermediate results in a
file or sending them over the network.  Programming langauges such as
Java, Ruby, Python, and R include built-in serialization support, but
these formats are tied to the specific programming language in use.
CSV files can be read and written by many applications and so are
often used for exporting tabular data.  However, CSV files have a
number of disadvantages, such as a limitation of exporting only
tabular datasets, lack of type-safety, inefficient text representation
and parsing, and abiguities in the format involving special
characters.  JSON is another widely supported format used mostly on
the web that removes many of these disadvantages, but it too suffers
from being too slow to parse and also does not provide strong typing
between integers and floating point.  Large numbers of JSON messages
would also be required to duplicate the field names with each message.

@@

@@

The protocol buffer project page contains a comprehensive
description of the language: \url{http://code.google.com/apis/protocolbuffers/docs/proto.html}

@@

\section{Dynamic use: Protocol Buffers and R}

\subsection{Importing proto files}

In contrast to the other languages (Java, C++, Python) that are officially
supported by Google, the implementation used by the \texttt{RProtoBuf}
package does not rely on the \texttt{protoc} compiler (with the exception of
the two functions discussed in the previous section). This means that no
initial step of statically compiling the proto file into C++ code that is
then accessed by R code is necessary. Instead, \texttt{proto} files are
parsed and processed \textsl{at runtime} by the protobuf C++ library---which
is much more appropriate for a dynamic language.

The \texttt{readProtoFiles} function allows importing \texttt{proto}
files in several ways.

Using the \texttt{file} argument, one can specify one or several file
paths that ought to be proto files.

<<>>=
proto.dir <- system.file( "proto", package = "RProtoBuf" )
proto.file <- file.path( proto.dir, "addressbook.proto" )
<<eval=FALSE>>=
readProtoFiles( proto.file )
@

With the \texttt{dir} argument, which is
ignored if the \texttt{file} is supplied, all files matching the
\texttt{.proto} extension will be imported.

<<>>=
dir( proto.dir, pattern = "\\.proto$", full.names = TRUE )
<<eval=FALSE>>=
readProtoFiles( dir = proto.dir )
@

Finally, with the
\texttt{package} argument (ignored if \texttt{file} or
\texttt{dir} is supplied), the function will import all \texttt{.proto}
files that are located in the \texttt{proto} sub-directory of the given
package. A typical use for this argument is in the \texttt{.onLoad}
function of a package.

<<eval=FALSE>>=
readProtoFiles( package = "RProtoBuf" )
@

Once the proto files are imported, all message descriptors are
are available in the R search path in the \texttt{RProtoBuf:DescriptorPool}
special environment. The underlying mechanism used here is
described in more detail in section~\ref{sec-lookup}.

@@

@@

The objects contained in the special environment are
descriptors for their associated message types. Descriptors will be
discussed in detail in another part of this document, but for the
purpose of this section, descriptors are just used with the \texttt{new}
function to create messages.

<<>>=
p <- new( tutorial.Person, name = "Romain", id = 1 )
@

@@

However, the main focus of protocol buffer messages is
efficiency. Therefore, messages are transported as a sequence
of bytes. The \texttt{serialize} method is implemented for
protocol buffer messages to serialize a message into the sequence of
bytes (raw vector in R speech) that represents the message.

@@

<<>>=
tf1 <- tempfile()
tf1
serialize( p, tf1 )
readBin( tf1, raw(0), 500 )
@

@@

The \texttt{RProtoBuf} package defines the \texttt{read}
function to read messages from files, raw vector (the message payload)
and arbitrary binary connections.

<<>>=
args( read )
@

@@

<<>>=
message <- read( tutorial.Person, tf1 )
writeLines( as.character( message ) )
@

@@

\section{Basic Abstractions: Messages, Descriptors, and
  DescriptorPools}

The three basic abstractions of \CRANpkg{RProtoBuf} are Messages,
which encapsulate a data structure, Descriptors, which define the
schema used by one or more messages, and DescriptorPools, which
provide access to descriptors.

@@

>>>>

Modern data collection and analysis pipelines are increasingly being
built using collections of components to better manage software
complexity through reusability, modularity, and fault
isolation \citep{Wegiel:2010:CTT:1932682.1869479}.  Different
programming languages are often used for the different phases of data
analysis -- collection, cleaning, analysis, post-processing, and
presentation in order to take advantage of the unique combination of 
performance, speed of development, and library support offered by
different environments.  Each stage of the data
analysis pipeline may involve storing intermediate results in a
file or sending them over the network.  Programming langauges such as
Java, Ruby, Python, and R include built-in serialization support, but
these formats are tied to the specific programming language in use.
% TODO(ms): and they often don't support versioning among other faults.
CSV files can be read and written by many applications and so are
often used for exporting tabular data.  However, CSV files have a
number of disadvantages, such as a limitation of exporting only
tabular datasets, lack of type-safety, inefficient text representation
and parsing, and abiguities in the format involving special
characters.  JSON is another widely supported format used mostly on
the web that removes many of these disadvantages, but it too suffers
from being too slow to parse and also does not provide strong typing
between integers and floating point.  Large numbers of JSON messages
would also be required to duplicate the field names with each message.

@@

% TODO(ms) Also talk about versioning and why its useful.

@@

\section{Basic Usage: Messages and Descriptors}

The two fundamental building blocks of Protocol Buffers are Messages
and Descriptors.  Messages provide a common abstract encapsulation of
structured data fields of the type specified in a Message Descriptor.
Message Descriptors are defined in \texttt{.proto} files and define a
schema for a particular named class of messages.  This separation
between schema and the message objects is in contrast to
more verbose formats like JSON, and when combined with the efficient
binary representation of any Message object explains a large part of
the performance and storage-space advantage offered by Protocol
Buffers. TODO(ms): we already said some of this above.  clean up.

Table~\ref{tab:proto} shows an example \texttt{.proto} file which
defines the \texttt{tutorial.Person} type.  The R code in the right
column shows an example of creating a new message of this type and
populating its fields.

% lifted from protobuf page:
%With Protocol Buffers you define how you want your data to be
%structured once, and then you can read or write structured data to and
%from a variety of data streams using a variety of different
%languages.  The definition

@@

\subsection{Importing Message Descriptors from \texttt{.proto} files}

%The three basic abstractions of \CRANpkg{RProtoBuf} are Messages,
%which encapsulate a data structure, Descriptors, which define the
%schema used by one or more messages, and DescriptorPools, which
%provide access to descriptors.

Before we can create a new Protocol Buffer Message or parse a
serialized stream of bytes as a Message, we must read in the message
type specification from a \texttt{.proto} file.

New \texttt{.proto} files are imported with the \code{readProtoFiles}
function, which can import a single file, all files in a directory, or
all \texttt{.proto} files provided by another R package.

The \texttt{.proto} file syntax for defining the structure of protocol
buffer data is described comprehensively on Google Code:
\url{http://code.google.com/apis/protocolbuffers/docs/proto.html}.

Once the proto files are imported, all message descriptors are
are available in the R search path in the \texttt{RProtoBuf:DescriptorPool}
special environment. The underlying mechanism used here is
described in more detail in Section~\ref{sec-lookup}.

@@

%\subsection{Importing proto files}
%In contrast to the other languages (Java, C++, Python) that are officially
%supported by Google, the implementation used by the \texttt{RProtoBuf}
%package does not rely on the \texttt{protoc} compiler (with the exception of
%the two functions discussed in the previous section). This means that no
%initial step of statically compiling the proto file into C++ code that is
%then accessed by R code is necessary. Instead, \texttt{proto} files are
%parsed and processed \textsl{at runtime} by the protobuf C++ library---which
%is much more appropriate for a dynamic language.

@@

New messages are created with the \texttt{new} function which accepts
a Message Descriptor and optionally a list of ``name = value'' pairs
to set in the message.
%The objects contained in the special environment are
%descriptors for their associated message types. Descriptors will be
%discussed in detail in another part of this document, but for the
%purpose of this section, descriptors are just used with the \texttt{new}
%function to create messages.

<<>>=
p1 <- new( tutorial.Person )
p <- new( tutorial.Person, name = "Romain", id = 1 )
@

@@

However, the main focus of protocol buffer messages is
efficiency. Therefore, messages are transported as a sequence
of bytes. The \texttt{serialize} method is implemented for
protocol buffer messages to serialize a message into a sequence of
bytes that represents the message.
%(raw vector in R speech) that represents the message.

@@

<<>>=
tf1 <- tempfile()
serialize( p, tf1 )
readBin( tf1, raw(0), 500 )
@

@@

The \texttt{RProtoBuf} package defines the \texttt{read} and
\texttt{readASCII} functions to read messages from files, raw vectors,
or arbitrary connections.  \texttt{read} expects to read the message
payload from binary files or connections and \texttt{readASCII} parses
the human-readable ASCII output that is created with
\code{as.character}.

@@

<<>>=
msg <- read( tutorial.Person, tf1 )
writeLines( as.character( msg ) )
@

@@

@@

\begin{table}[h]
\centering
\begin{small}
\begin{tabular}{l|l}
\textbf{Method} & \textbf{Description} \\
\hline
\hline
\texttt{has} & Indicates if a message has a given field.   \\
\texttt{clone} & Creates a clone of the message \\
\texttt{isInitialized} & Indicates if a message has all its required fields set\\
\texttt{serialize} & serialize a message to a file, binary connection, or raw vector\\
\texttt{clear} & Clear one or several fields of a message, or the entire message\\
\texttt{size} & The number of elements in a message field\\
\texttt{bytesize} & The number of bytes the message would take once serialized\\
\hline
\texttt{swap} & swap elements of a repeated field of a message\\
\texttt{set} & set elements of a repeated field\\
\texttt{fetch} & fetch elements of a repeated field\\
\texttt{setExtension} & set an extension of a message\\
\texttt{getExtension} & get the value of an extension of a message\\
\texttt{add} & add elements to a repeated field \\
\hline
\texttt{str} & the R structure of the message\\
\texttt{as.character} & character representation of a message\\
\texttt{toString} & character representation of a message (same as \texttt{as.character}) \\
\texttt{as.list} & converts message to a named R list\\
\texttt{update} & updates several fields of a message at once\\
\texttt{descriptor} & get the descriptor of the message type of this message\\
\texttt{fileDescriptor} & get the file descriptor of this message's descriptor\\
\hline
\end{tabular}
\end{small}
\caption{\label{Message-methods-table}Description of methods for the \texttt{Message} S4 class}
\end{table}

\subsection{Descriptors}

Message descriptors are represented in R with the
\emph{Descriptor} S4 class. The class contains
the slots \texttt{pointer} and \texttt{type} :

\begin{table}[h]
\centering
\begin{tabular}{|cp{10cm}|}
\hline
\textbf{slot} & \textbf{description} \\
\hline
\texttt{pointer} & external pointer to the \texttt{Descriptor} object of the C++ proto library. Documentation for the
\texttt{Descriptor} class is available from the protocol buffer project page:
\url{http://code.google.com/apis/protocolbuffers/docs/reference/cpp/google.protobuf.descriptor.html#Descriptor} \\
\hline
\texttt{type} & fully qualified path of the message type. \\
\hline
\end{tabular}
\caption{\label{Descriptor-class-table}Description of slots for the \texttt{Descriptor} S4 class}
\end{table}

Similarly to messages, the \verb|$| operator can be used to extract
information from the descriptor, or invoke pseuso-methods.

\subsubsection{Extracting descriptors}

The \verb|$| operator, when used on a descriptor object retrieves
descriptors that are contained in the descriptor.

This can be a field descriptor (see section~\ref{subsec-field-descriptor} ),
an enum descriptor (see section~\ref{subsec-enum-descriptor}) or a descriptor
for a nested type

<<>>=
# field descriptor
tutorial.Person$email

# enum descriptor
tutorial.Person$PhoneType

# nested type descriptor
tutorial.Person$PhoneNumber
# same as
tutorial.Person.PhoneNumber
@

\begin{table}[h]
\centering
\begin{small}
\begin{tabular}{l|l}
\textbf{Method} & \textbf{Description} \\
\hline
\hline
\texttt{new} & Creates a prototype of a message described by this descriptor.\\
\texttt{read} & Reads a message from a file or binary connection.\\
\texttt{readASCII} & Read a message in ASCII format from a file or
text connection.\\
\hline
\texttt{name} & Retrieve the name of the message type associated with
this descriptor.\\
\texttt{as.character} & character representation of a descriptor\\
\texttt{toString} & character representation of a descriptor (same as \texttt{as.character}) \\
\hline
\texttt{fileDescriptor} & Retrieve the file descriptor of this
descriptor.\\
\texttt{containing\_type} & Retrieve the descriptor describing the message type containing this descriptor.\\
\texttt{field\_count} & Return the number of fields in this descriptor.\\
\texttt{field} & Return the descriptor for the specified field in this descriptor.\\
\texttt{nested\_type\_count} & The number of nested types in this descriptor.\\
\texttt{nested\_type} & Return the descriptor for the specified nested
type in this descriptor.\\
\texttt{enum\_type\_count} & The number of enum types in this descriptor.\\
\texttt{enum\_type} & Return the descriptor for the specified enum
type in this descriptor.\\
\hline
\end{tabular}
\end{small}
\caption{\label{Descriptor-methods-table}Description of methods for the \texttt{Descriptor} S4 class}
\end{table}
